---
title: "Reviewing content of Sept 30 meeting of BST 2026"
keywords: "Probability"
date: 2026-09-30
format: html
---
# Finite and infinite sets

There is an intimate connection between probability modeling
in data science and the mathematics of sets.  We distinguished

- Finite sets: the elements can be mapped one-to-one to the set $\{1, \ldots, n\}$ for
a fixed and specified number $n$.
- Countably infinite sets: not finite, and elements can be mapped one-to-one to the natural numbers.  
- Uncountably infinite sets: not finite, but there is no one-to-one mapping to the
natural numbers.  The existence of such sets was 
demonstrated by [Georg Cantor](https://en.wikipedia.org/wiki/Georg_Cantor).

# Discrete and continuous probability models

## Discrete models

We have been informal in our treatment of random outcomes in chance
situations, such as drawing a card from a well-shuffed deck.  The [notes from
September 27](https://vjcitn.github.io/2026vjc/slides/prob/20b-chisq-cards.html#a-model-for-fairness)
provide some formalism.  We focused on the concepts of

- defining the index sets for components of experiments (e.g., let $i = 1, \ldots, 52$ denote the
different card faces available, let $j = 1, \ldots , 1000$ denote the number of draws)
- clarifying assumptions of independence
- recalling the form of the chi-squared statistic
$$
\sum_{i=1}^K \frac{(O_i-E_i)^2}{E_i}
$$
where $O_i$ and $E_i$ are observed and expected counts for events of category $i$.
When the contributions to the observed counts are independent and
all categories are equiprobable, this sum
can be referred to the chi-squared distribution with $K-1$ degrees of freedom.

We also used the murders dataset to form a contingency table and test the
goodness of fit of a model prescribing equal risk of death by homicide
across all states and D.C.

## Continuous models

Measures like height and blood pressure are typically regarded as real numbers,
and we use double-precision floating point numbers to compute on these.  The fact
that our computations are based on data values that are recorded with limited precision
does not prevent us from using continuous probability models for such outcomes.

We reviewed relevant content in the Intro to Probability slide deck and corrected
the form of the Gaussian density in [this slide](https://vjcitn.github.io/2026vjc/slides/prob/20c-intro-to-prob.html#/probability-density-function-3).

# Exploring the CDC waterborne/food-related illness API response
```
library(httr)
z = GET("https://data.cdc.gov/api/v3/views/5xkq-dg7x/query.json")
```
We used `zc = content(z)` but then saved the result of that to
`waterborne.rda`.  Note that the API content was 29MB, but
the compressed data in RDA format is 2.9MB in size.

We began "getting a sense"
of the CDC data on waterborne and food disease outbreaks.

This means looking at a list of lists, and learning the
names for concepts handled in one of the lists.  It may
not be the case that all elements of the top level
list have identical structure.  How
would you find that out?

Here we go, without addressing that question.
First we learn about the overall list length.
```{r lk1}
load("waterborne.rda")
zc = waterborne
class(zc)
length(zc)
```
Then we probe one element.
```{r lk1el}
zc[[1]]
```

We can use an applicative functional
to _try to_ grab date data from each element.
`sapply` is nice because it will simplify to
a vector or array if possible.
```{r lksapp}
years = sapply( zc, function(x) x$year)
head(years)
length(years)
```

To get a sense of history of the data 
collected, a histogram of year values would be good.
But it turns out the data are character!
We have to coerce.
```{r lkhist}
try(hist(years))
hist(as.numeric(years))
```

Mode of disease incidence is of interest:
```{r lkmode}
modes = sapply( zc, function(x) x$primary_mode)
table(modes)
```


Here we realize that some elements have
no information for some fields.
```{r chkset}
settings = sapply( zc, function(x) x$setting)
head(settings)
table(sapply(settings,length))
```
So we decide to create a more intelligent
collection function.

```{r domoret}
etio = sapply( zc, function(x) x$etiology)
table(sapply(etio,length))
etio = sapply( zc, function(x) { tmp = x$etiology; if(length(tmp)==0) return(NA); return(tmp)} )
table(etio) |> sort() |> tail(20)
```
OK, we are satisfied that we can get
interesting information from the list.
But let's try to have AI analyze the
structure and produce an app that
simplifies our exploration of the resource.

We wrote out a subset of the data so as not
to overwhelm the chatbot.  But I made an error, using [5000] instead of [1:5000]...
```
library(jsonlite)
write_json(zc[1:5000], "cdc.json")
```

While it was dramatic to ask the chatbot to "create an app that we can
use to explore the data", it would have been more reasonable to start
with creation of a tidy data table ...
