---
title: "Distributions, PRNG, moments, lm"
author: "Vince Carey, instructor, Fall 2026 edition"
keywords: "Probability"
date: 2026-10-07
---

# Prologue from first day of course

The course includes activities in three main domains:

- surveying *data science and statistics* as foundations of basic integrity and efficiency of research
- introducing and using *tools for implementing data science methods* for research
- critically exploring opportunities afforded by *large language models and agentic AI* for automation of repetitive detailed tasks

A significant concern with AI adoption in education is the risk of "de-skilling": students need to
be able to 

- *guide* agentic processes to accomplish clear goals 
- *verify* success and supply solutions in case of agent failure
- *propose paths forward* after success is achieved

# The uniform distribution in the unit interval

We use $U(0,1)$ to denote a probability model for real numbers
$0 \leq x \leq 1$.  Under this model, the probability that a
random quantity lies in the interval $[a, b]$, $0 \leq a < b \leq 1$
is $b-a$.  Thus if $X \sim U(0,1)$ the probability that $X$
lies in $(0.40,0.55)$ is $0.15$.

*Exercise:* Use R to reproducibly verify this claim.  The function `runif` and
logical and comparison and averaging operations may be used.

*Exercise:* By default `runif` produces points with "random" locations
on the unit interval.  How can you produce points $(x,y)$ with "random"
locations on the unit square?  How would you test for adequacy of
this procedure?

# Random number generation

Our approach to exploring randomness, probability, and statistical
models is heavily based on _simulation_.  The two parts of simulation
studies are the experimental setup and the introduction of random
elements to produce artificial experimental outcomes.  How do computers
produce random events?

In general they do not.  Procedures that produce, deterministically,
"unpredictable, random-seeming series" have been devised.  An acronym
in use is PRNG: pseudo-random number generation.  Statistical
testing is used to assess procedures according to their adequacy in
resembling truly random processes.

## What we want

We want a sequence of double precision numbers that "appear" random.  We
use $x_1, \ldots$ to represent the sequence.

We want to be able to recompute any sequence generated previously.  Thus we must have
a representation of the state producing $x_1$, and our generator must
obey the setting of the initial state.

We want to be able to produce sequences that are "independent of one another".
"Hidden correlations" among sequences or sequence elements can lead to misleading
conclusions from simulation experiments.

## The linear congruential method

An important example is the "linear congruential method".  We use
the "mod" function that provides the remainder after integer
division.  It is denoted `%%` in R.

$$
x_{k+1} = (a\,x_k + c) \bmod m, \qquad k = 0, 1, 2, \ldots
$$

where

- $m > 0$ is the **modulus**,
- $a$ is the **multiplier**,
- $c$ is the **increment**,
- $x_0$ is the **seed**.

Every $x_k$ is an integer in $\{0, 1, \ldots, m-1\}$, a finite set, so the
period can be at most $m$. Dividing by $m$ gives numbers $u_k = x_k / m$ in $[0, 1)$.

Claude produced the following generator:
```{r}
lcg <- function(n, seed, a, c, m) {
  u <- numeric(n)
  x <- seed
  for (k in seq_len(n)) {
    x <- (a * x + c) %% m
    u[k] <- x / m
  }
  u
}
```
Effective choices of modulus, multiplier and increment has occupied
researchers for decades.  See the discussion of [Park and Miller (1988)](https://en.wikipedia.org/wiki/Lehmer_random_number_generator).  An implementation of the "minimum standard" generator is
```{r dominstd}
m31 <- 2^31
minstd <- function(n, seed) lcg(n, seed, a = 16807, c = 0, m = m31 - 1)  # Park & Miller (1988)
hist(minstd(2000, 1234))
```

*Exercise:* Define a chi-squared procedure for assessing the adequacy of this generator.

*Exercise:* What does the following result imply about the granularity of floating
point numbers produce by R's default PRNG?

```{r lkdup}
set.seed(1234)
tail(sort(table(runif(1e6))))
```

# Families of probability models

We have discussed "expectation" in an intuitive and informal way.
We have described probabilities a little more formally.  For now
our interpretation of probability is long-run relative frequency, and
"long run" means "established through an infinite number of trials".

We may use "model" or "distribution" to refer to the
conceptual frameworks used to describe random variables
or processes.  It is important to remember that any collection
of numbers forms a distribution, but a _probability distribution_
is a specific theoretical structure that may be used to help
understand aspects of collections of numbers we encounter in
scientific work.

## Bernoulli trials

We had the simplest possible simulation of card shuffles and draws.
We observed a single card per shuffle.  Fix attention
on a specific card face, say, 2 of hearts.  After a shuffle,
we draw that card from the top or we do not.  Conventionally
we use $X$ to denote the random variable that is 1 if the draw
is 2 of hearts, 0 if it is not.  The set $\{0, 1\}$ is the _sample
space_ for the experiment that examines the top card after a shuffle.
The probability model for this experiment obeys the
_Bernoulli distribution_ with success probability $p = 1/52$ for
a fair deck.

## The binomial distribution

Here we have a fixed number of trials that end in one of two possible outcomes.
As an example, the Saxony dataset in the vcdExtra package has counts
of numbers of male children in families with 12 children.  Let $X$ denote
the number of males, the model asserts that
$$
Pr(X = k) = \binom{n}{k} p^k(1-p)^{n-k}, k = 0, \ldots, n
$$
where $n = 12$ for the Saxony data, and $p$ is the probability that
a given child's sex is male.  Setting $p = 0.5$, we would have
```{r lksax,message=FALSE}
library(vcdExtra)
data(Saxony)
Saxony
```
as the observed data, and
```{r lkdbin}
dbinom(0:12, 12, .5)*6155
```
as the expected number of males under the specified binomial model.

*Exercise:*  The Saxony data represent counts of male sons
in families of size 12.  Produce a data.frame with one record
per individual that records family identifier and gender.
Use the data in this table to estimate the probability that a baby in
this collection of families was born male.

## The hypergeometric distribution

Consider the outcome of drawing five cards from a fair deck.
Our outcome is the number of hearts.  We are drawing without
replacement.  There are a total of $\binom{52}{5}$ ways
of selecting 5 cards, and the number of
ways of obtaining $k$ hearts is $\binom{13}{k} \times \binom{39}{5-k}$.
Thus
$$
Pr(X = k) = \frac{\binom{13}{k} \binom{39}{5-k}}{\binom{52}{5}}, k = 0, \ldots, 5.
$$

*Exercise:* Use our card-shuffling setup to estimate the probability
of getting a hand with exactly two hearts.  Assess the accuracy
of your estimate using dhyper.

## The Poisson distribution

A series representation of the antilog function $\exp(x)$ is
$$
e^x = \sum_{k=0}^\infty \frac{x^k}{k!}.
$$
Thus we can define a probability mass function, that of the Poisson distribution, as
$$
Pr(X = k) = \frac{e^{-\mu} \mu^k}{k!}, k = 0, 1, 2, \ldots.
$$
The values of $k$ are all the non-negative integers.  This is an example of
a countably infinite sample space.  The parameter $\mu$ is the average value
of $X$.  

It is typical to use the Poisson distribution as a first approach
to modeling rates of rare events.  In the case of the murders dataset,
let $\lambda$ denote the risk per million individuals of being
a victim of homicide.
Then we can write $\mu = \lambda N/10^6$ as the average count for a state of
size $N$ and estimate estimate $\lambda$ as the mean number of murders per million
population.

```{r dopoimurd, message=FALSE}
library(dslabs)
data(murders)
lhat = sum(murders$total)/sum(murders$population)*1e6
round(lhat,3)
pd = data.frame(pred=lhat*murders$population/1e6, obs=murders$total)
plot(pd$obs, pd$pred, xlab="observed count", ylab="model-based prediction")
```

Since we are seeing variation over orders of magnitude, log transformation is in order.
```{r dolog}
plot(pd$obs, pd$pred, xlab="observed count", ylab="model-based prediction", log="xy")
```
The Poisson model prescribes probabilities for counts of murders per million inhabitants.
```{r lkdenss}
plot(0:50, dpois(0:50, lhat), main="Probabilities for common rate of murder per million population, all states")
```

Later we consider how to evaluate adequacy of such models in application to specific data.

## The Gaussian distribution

We have seen the expression for the Gaussian density in the slides.
$$
f(x) = \frac{1}{\sqrt{2 \pi \sigma^2}}\exp\{ - \frac{(x - \mu)^2}{2 \sigma^2} \}
$$
Parameters $\mu$ and $\sigma^2$ determine location and dispersion of
the distribution.

## Working with distributions in R

The d-, p-, q-, r- functions are available for a wide variety of probability models.
Examples for Poisson:

- `dpois(x, mu)` computes the probability mass function for the Poisson model with mean value `mu`
- `ppois(x, mu)` computes the probability that data from a Poisson model with mean value `mu`
take values less than or equal to x
- `qpois(p, mu)` computes the value below which `p` percent of data from a Poisson model with
mean value `mu` will be found
- `rpois(n, mu)` produces a vector with `n` random values representing realizations of the
Poisson model with mean value `mu`

Try them out.  First, with a discrete distribution.
```{r trypo}
set.seed(1234)
x = rpois(1000, 4.5)
mean(x)
(pat3 = ppois(3.2, 4.5))
(fr = mean(x <= 3.2))
qpois(fr, 4.5)
(pat1 = ppois(1.8, 4.5))
(fr = mean(x <= 1.8))
qpois(fr, 4.5)
```
Now with a continuous distribution.
```{r tryga}
set.seed(1234)
x = rnorm(10000, 4.5)
mean(x)
(pat3 = pnorm(3.2, 4.5))
(fr = mean(x <= 3.2))
qnorm(fr, 4.5)
(pat1 = pnorm(1.8, 4.5))
(fr = mean(x <= 1.8))
qnorm(fr, 4.5)
```

Here are some function traces to know and understand.  `ecdf` is particularly
interesting for its design, returning a function.
```{r lktraces}
par(mfrow=c(2,2))
curve(dnorm, -4,4)
curve(pnorm, -4,4)
curve(qnorm, .001, .999)
arrows(.25,-3,.25,qnorm(.25),length=.1)
arrows(.25,qnorm(.25),0,qnorm(.25),length=.1)
x = rnorm(500)
plot(ecdf(x))
```



# Model parameters and sample statistics

## Definitions

We have been very informal in our discussion of expected values.  There
are two important expectation concepts to get clear on.

For a random variable $X$ with discrete distribution on sample space $S$ and probability mass function $p$,
the _expectation_ is the quantity
$$
E(X) = \sum\limits_{x \in S} xp(x)
$$
which we sometimes denote $\mu$.

The _variance_ is the quantity
$$
Var(X) = E((X - E(X))^2) = \sum\limits_{x \in S} (x - \mu)^2 p(x).
$$

For continuous distributions like the Gaussian or Chi-squared, the density function
is used:
$$
E(X) = \int\limits_{x \in S} xf(x)dx;
$$
$$
Var(X) =  \int\limits_{x \in S} (x - \mu)^2 f(x)dx.
$$

The sample average is often used to estimate the expected value of
a random variable.  For $n$ observations it is defined as
$$
\bar{x} = \frac{x_1 + x_2 + \cdots + x_n}{n}.
$$
The function `mean` in R computes this for any vector of numbers.

The sample variance is defined as
$$
\hat{s}^2 = \frac{\sum_i (x_i - \bar{x})^2}{n-1}
$$
and this is computed by `var` in R.  The denominator
used reflects the degree of freedom lost by estimating $\bar{x}$
from the data.  

Finally, we introduce a measure of bivariate association: covariance.
$$
Cov(X, Y) = E(XY) - E(X)E(Y).
$$
Note that $Cov(X,X) = Var(X)$.

The correlation coefficient is the covariance divided by the product of standard deviations of $X$ and $Y$.
$$
Corr(X, Y) = \frac{Cov(X, Y)}{\sqrt{Var(X) \cdot Var(Y)}}
$$

## Examples

```{r doit,message=FALSE}
library(NHANES)
library(dplyr)
library(ggplot2)
data(NHANES)
NHANES|> filter(BPDiaAve>0) |> select(BPDiaAve, Sex) |> 
    summarise(m=mean(BPDiaAve, na.rm=TRUE), s=sd(BPDiaAve, na.rm=TRUE), .by=Sex)
ggplot(NHANES |> filter(BPDiaAve>0), aes(x=BPDiaAve)) + 
    geom_area(stat="density") + facet_wrap(Sex~., ncol=1)
```
The quantitative differences in mean and dispersion are not
readily perceived in the juxtaposition of densities.

For the bivariate measure, we look at SBP and DBP jointly.
```{r lkcov}
cov(NHANES$BPSysAve, NHANES$BPDiaAve, use="complete.obs")
cor(NHANES$BPSysAve, NHANES$BPDiaAve, use="complete.obs")
```
The covariance matrix for collection of random variables
has pairwise covariances off the diagonal, and marginal
variances on the diagonal.  The correlation matrix has
1 on diagonal, and pairwise correlations off the diagonal.
```{r lkcovmat}
cov(cbind(NHANES$BPSysAve, NHANES$BPDiaAve), use="complete.obs")
cor(cbind(NHANES$BPSysAve, NHANES$BPDiaAve), use="complete.obs")
```
We can visualize with hexagon bins to avoid overplotting.
```{r lkjhex}
ggplot(NHANES |> filter(BPDiaAve>0), aes(x=BPDiaAve, y=BPSysAve)) + 
    geom_hex() + facet_wrap(Sex~., ncol=2)
```

# "Compact estimation" from tables

## A call to `lm`

Let's focus on the 2011-2012 subset from NHANES.

```{r lksub,message=FALSE}
nh11 = NHANES |> filter(SurveyYr == "2011_12")
library(DT)
datatable(nh11[1:100, 1:6])
```

Scroll through the table.  Note that there are repeated
values of ID!  This is a consequence of the "resampling"
process that generated `NHANES` from `NHANESraw`
in the package.  Individuals in this dataset
represent groups of individuals with similar demographic characteristics.
Multiple "copies" of individuals will be present for those
individuals with highly prevalent configurations of
demographic variables.  We are going to ignore this
feature of the data in the remainder of this document.
We will compute statistics on variables in this dataset
as if they are of interest in their own right, collected
on independent individuals, with no
specific epidemiological inference proposed.

```{r lkmo,eval=FALSE}
ss = split(nh11, nh11$ID)
nr = sapply(ss,nrow)
o = order(nr, decreasing=TRUE)
datatable(rbind(ss[[o[1]]], ss[[o[2]]]))
```

The following dplyr code assembles a table of
means of systolic blood pressure in strata formed
by decade of age.

```{r dodec}
nh11 |> summarise(m=mean(BPSysAve,na.rm=TRUE), .by=AgeDecade)
```
The same mean values (with the exception of
the one for missing AgeDecade) are recovered by using the
`lm` function with specific arguments for formula and data.
```{r dolm1}
lm(BPSysAve~AgeDecade-1, data=nh11)$coef
```

Question: why are the values with missing age ignored in
the call to `lm`?

Question: how can these values be used to obtain an additional
estimated mean BPSysAve?

Question: what is the role of `$coef` in our use of `lm`?

## Wilkinson-Rogers (1973) formula notation

### Basic form

```r
response ~ terms
```

- Left of `~`: response. It can be transformed (`log(y)`)
- Right of `~`: a sum of terms, each expanding to one or more design-matrix columns.
- An intercept is included by default.

### Operators

| Operator | Meaning | Expansion |
|---|---|---|
| `+` | add a term | `a + b` → main effects of a, b |
| `:` | interaction only | `a:b` → the a×b interaction |
| `*` | crossing | `a*b` = `a + b + a:b` |
| `/` | nesting | `a/b` = `a + a:b` (b nested within a) |
| `%in%` | nested within | `b %in% a` = `a:b`-style term |
| `^n` | all interactions up to order n | `(a+b+c)^2` = `a+b+c+a:b+a:c+b:c` |
| `-` | remove a term | `a*b - a:b` = `a + b` |
| `-1` / `+0` | no intercept | `y ~ x - 1` |
| `1` | intercept (explicit) | `y ~ 1` is the null model |
| `.` | all other columns in `data` | `y ~ .` |
| `I()` | arithmetic as-is | `I(x^2)`, `I(a+b)` |

Inside a formula, `^`, `*`, `+` and `-` are formula operators. Wrap in `I()` when you mean ordinary arithmetic.

### Terms can be functions

`log(x)`, `poly(x, 2)`, `splines::ns(x, 3)`, `factor(g)`, `offset(log(t))`.

