---
title: "R Basics"
keywords: "R"
date: 2026-09-14
format:
  revealjs:
    theme: night
execute: 
  echo: true
---


## Packages

* R has a *base* installation and then tens of thousands of add-on software packages that can be obtained from CRAN

* Use `install.packages` to install the **dslabs** package from [CRAN](https://cran.r-project.org/)

* Try out the following functions: `sessionInfo`, `installed.packages`

---

## Prebuilt functions

* Much of what we do in R uses functions from either base R or installed packages.

* Many are included in automatically loaded packages: `stats`, `graphics`, `grDevices`, `utils`, `datasets`, `methods`.

* This subset of the R universe is referred to as _R base_.

* Very popular packages not included in R base: `ggplot2`, `dplyr`, `tidyr`, and `data.table`.

* It is easy to write your own functions and packages

---

:::{.callout-important}
For problem set 2 you can only use R base.
:::

---


## Base R functions

* Example functions that we will use today: `ls`, `rm`, `library`, `search`, `factor`, `list`, `exists`, `str`, `typeof`, and `class`.

* You can see the raw code for a function by typing its name without the parentheses.

    * type `ls` on your console to see an example.

---

## Help system

* In R you can use `?` or `help` to learn more about functions.

* You can learn about function using

```r
help("ls")
```

or 

```r
?ls
```
* many packages provide *vignettes* that are like how-to manuals that can show how the functions in the package are meant to be used to create different analyses

---

## The workspace

* Define a variable.

```{r}
a <- 2
```

* Use `ls` to see if it's there. Also take a look at the **Environment** tab in RStudio.

```{r}
ls()
```

* Use `rm` to remove the variable you defined.

```{r}
rm(a)
```

---

## The workspace{.smaller}

* each time you start R you will get a new workspace that does not have any variables or libraries loaded

* when you quit R you will be asked if you want to save the workspace

  * you will probably always be better off saying **NO**

* if you do save a workspace it will be saved as a hidden file (Unix lecture) and whenever R is started in a directory with a saved workspace then that workspace will be re-instantiated and used

  * this might be helpful if you are working on a long complex analysis with large data
  
* mostly it is much easier to just use a markdown document to detail your steps and rerun them on a clean workspace

  * this can also help find errors in your analysis
  
--- 

## The Workspace

* just like Unix R uses a search path to find functions that you can evaluate

```{r search}
search()
```

---

## Variable name convention {.smaller}

* A nice convention to follow is to use **meaningful** words that describe what is stored, **only lower** case, and **underscores** as a substitute for spaces. 

* R and RStudio both provide autocomplete capabilities so you don't need to type the whole name     
    * this makes it easier to use longer, more descriptive variable names

* It is highly recommended that you not use the period `.` in variable names, there are situations where R treats it differently and those can cause unintended actions

* For more we recommend [this guide](http://adv-r.had.co.nz/Style.html).

--- 

## Data types{.smaller}

The main data types in R are:

* One dimensional vectors: double, integer, logical, complex, characters.

* integer and double are both numeric

* Factors

* Lists: this includes data frames.

* Arrays: Matrices are the most widely used.

* Date and time

* tibble

* S4 objects

---

## Data types

* Many errors in R come from confusing data types. 

* `str` stands for structure, gives us information about an object.

* `typeof` gives you the basic data type of the object. It reveals the lower-level, more fundamental type of an object in R's memory.

* `class` This function returns the class attribute of an object. The class of an object is essentially `type_of` at a higher, often user-facing level.

---

## Data types{.smaller}

Let's see some example:

```{r}
library(dslabs)
typeof(murders)
class(murders)
str(murders)
dim(murders)
```

---

## Data frames{.smaller}

* Date frames are the most common class used in data analysis. 

* Data frames are like a matrix, but where the columns can have different types.

* Usually, rows represents observations and columns variables. 

* you can index them like you would a matrix,  `x[i, j]` refers to the element in row `i` column `j`

* You can see part of the content like this

```{r}
head(murders)
```


---

## Data frames

* and you can use `View` to open a spreadsheet-like interface to see the entire data.frame.

```{r}
#| eval: false
View(murders)
```

* This is more effective in RStudio because it has an integrated viewer. 

---

## Data frames{.smaller}

* A very common operation is adding columns like this:

```{r}
murders$pop_rank <- rank(murders$population)
head(murders)
```


---

## Data frames

* Note that we used the `$`.

* This is called an `accessor` because it lets us access columns.

```{r}
murders$population
```

* More generally: `$` can be used to access named components of a list.

---

## Data frames

* One way R confuses beginners is by having  multiple ways of doing the same thing.

* For example you can access the 4th column in the following five different ways: 

```{r}
#| eval: false

murders$population
murders[, "population"]
murders[, 4]
murders[["population"]]
murders[[4]]
```

* In general, we recommend using the name rather than the number as adding or removing columns will change index values, but not names.

---

## with

*  `with` let's us use the column names as symbols to access the data.


* This is convenient to avoid typing the data frame name over and over:

```{r}
rate <- with(murders, total/population)
```

---

## with

* Note you can write entire code chunks by enclosing it in curly brackets:

```{r}
with(murders, {
   rate <- total/population
   rate <- round(rate*10^5)
   print(rate[1:5])
})
```

---

## Vectors

* The columns of data frames are an example of one dimensional (atomic) vectors.

* An atomic vector is a vector where every element must be the same type.


```{r}
length(murders$population)
typeof(murders$population)
```

---

## Vectors{.smaller}


* Often we have to create vectors.  

* The concatenate function `c` is the most basic way used to create vectors:

```{r}
x <- c("b", "s", "t", " ", "2", "6", "0")
```

* We access the elements using `[]`

```{r}
x[5]
typeof(x[5])
x[12]
```
* **NOTE** R does not do array bounds checking...it silently pads with missing values

___

## Sequences {.smaller}

* Sequences are a common example of vectors we generate.

```{r}
seq(1, 10)
seq(1, 9, by=2)
```

* When you want a sequence that increases by 1 you can use the colon `:`

```{r}
1:10
3.3:7.8
4:-1
```

___

## Sequences{.smaller}

* A useful function to quickly generate a sequence the same length as a vector `x` is `seq_along(x)` and **NOT** `1:length(x)`

```{r}
x <- c("b", "s", "t", " ", "2", "6", "0")
seq_along(x)
```

* A reason to use this is to loop through entries:

```{r}
for (i in seq_along(x)) {
  cat(toupper(x[i]))
}
```

* But if the length of `x` is zero, then using `1:length(x)` does not work for this loop
```{r}
w = x[x=="W"]
1:length(w)
seq_along(w)
```

---  

## Vector types and coercion{.smaller}

* One dimensional vectors: double, integer, logical, complex, characters and numeric

* Each basic type has its own version of NA (a missing value)

* testing for types: `is.TYPE`, 
* coercing :   `as.TYPE`, will result in NA if it is not possible

```{r coercion}
is.numeric("a")
is.double(1L)
as.double("6")
as.numeric(x) + 3
typeof(1:10)
```

---

## Coercing

* When you do something inconsistent with data types, R tries to figure out what you mean and change it accordingly.

* We call this _coercing_.

* R does not return an error and in some cases does not return a warning either.

* This can cause confusion and unnoticed errors. 

* So it's important to understand how and when it happens. 

--- 

## Vector types and coercion{.smaller}

* coercion is automatically performed when it is possible

* `TRUE` coerces to 1 and `FALSE` to 0 
*  but any non-zero integer coerces to TRUE, only 0 coerces to FALSE
* `as.logical` converts 0 and only 0 to `FALSE`, everything else to `TRUE`
* the character string "NA" is not a missing value

```{r}
typeof(1:10 + 0.1)
typeof(TRUE+1)
as.character(TRUE)
as.numeric(TRUE)
as.logical(1)
as.logical(.5)
```

---

## Coercing {.smaller}

* Here are some examples:

```{r}
typeof(1L)
typeof(1)
typeof(1 + 1L)
c("a", 1, 2)
TRUE + FALSE
factor("a") == "a"
identical(factor("a"), "a")
```

--- 

## Coercing 

* When R can't figure out how to coerce, rather an error it returns an NA:

```{r}
as.numeric("a")
```

* Note that including `NA`s in arithmetical operations usually returns an `NA`.

```{r}
1 + 2 + NA
```

--- 

## Coercing

* You can coerce explicitly 

* Most coercion functions start with `as.`

* Here is an example.

```{r}
x <- factor(c("a","b","b","c"))
as.character(x)
as.numeric(x)
```

--- 

## Coercing

* The `readr` package provides some tools for trying to par 
```{r}
x <- c("12323", "12,323")
as.numeric(x)

library(readr)
parse_guess(x)
```

--- 

## Factors

* One key distinction between data types you need to understad is the difference between factors and characters.

* The `murder` dataset has examples of both.

```{r}
class(murders$state)
class(murders$region)
```

* Why do you think this is?

---  

## Factors {.smaller}

* A factor is a good representation for a variable, that has a fixed set of non-numeric values
    * ex: Sex has `Male` and `Female`

* It is usually not a good representation for variables that have lots of levels (like state names in the murders dataset)

* Internally a factor is stored as the unique set of labels (called levels) and an integer vector with values in 1 to `length(levels)`
  * if the `i`th entry is `k` then that corresponds to the `k`th element of the levels

* In statistics we refer to this as categorical data, where all of the individuals are mapped into a relatively small number of categories.

* Usually order does not matter, but if it does you can also have ordered factors.

```{r}
x <- murders$region
levels(x)
```

---

## Setting Levels

* you can set up the levels as you would like, when creating a factor

* if you do not set them up, then they will be created in lexicographic order (in the locale you are using)

```{r}
#| code-line-numbers: false

x = sample(c("Male", "Female"), 50, replace =TRUE)
y1 = factor(x, levels=c("Male", "Female"))
y2 = factor(x, levels = c("Female", "Male"))
y1[1:10]
y2[1:10]

```
---  

## Categories based on strata

* In data analysis we often want to stratify continuous variables into categories.

* The function `cut` helps us do this:

* In this case there may be a reason to think of using ordered factors.


```{r}
age <- c(5, 93, 18, 102, 14, 22, 45, 65, 67, 25, 30, 16, 21)
cut(age, c(0, 11, 27, 43, 59, 78, 96, Inf))
```


---  

## Categories based on strata

* We can assign more meaningful level names:

```{r}
age <- c(5, 93, 18, 102, 14, 22, 45, 65, 67, 25, 30, 16, 21)
cut(age, c(0, 11, 27, 43, 59, 78, 96, Inf), 
    labels = c("Alpha", "Zoomer", "Millennial", "X", "Boomer", "Silent", "Greatest"))
```

---  

## Changing levels

* This is often needed for ordinal data because R defaults to alphabetical order

* Or as noted you may want to make use of `ordered` factors

```{r}
gen <- factor(c("Alpha", "Zoomer", "Millennial"))
levels(gen)
```

* You can change this with the `levels` argument:

```{r}
gen <- factor(gen, levels = c("Alpha", "Zoomer", "Millennial", "X", "Boomer", "Silent", "Greatest"))
levels(gen)
```


---  

## Changing levels

* A common reason we want to change levels is to assure R is aware which is the reference strata.

* This is important for linear models because the first level is assumed to be the reference.

```{r}
x <- factor(c("no drug", "drug 1", "drug 2"))
levels(x)
x <- relevel(x, ref = "no drug")
levels(x)          
```

---  

## Changing levels

* We often want to order strata based on a summary statistic.

* This is common in data visualization.

* We can use `reorder` for this:

```{r}
x <- reorder(murders$region, murders$population, sum)
```


---  

## Factors

* Another reason we used factors is because they are stored more efficiently:

```{r}
x <- sample(murders$state[c(5,33,44)], 10^7, replace = TRUE)
y <- factor(x)
object.size(x)
object.size(y)
```

* An integer uses less memory than a character string (but it is a bit more complicated)

--- 

## Factors{visibility="hidden"}



Exercise: How can we make this go much faster?

```{r}
##FIXME:RG - not sure what this is supposed to show
system.time({levels(y) <- tolower(levels(y))})
```

--- 

## Factors can be confusing

* Try to make sense of this:

```{r}
x <- factor(c("3","2","1"), levels = c("3","2","1"))
as.numeric(x)

x[1]
levels(x[1])

table(x[1])
```

--- 

## Factors can be confusing

* Avoid keeping extra levels with `droplevels`:

```{r}
z <- x[1]
z <- droplevels(z)
```

* But note what happens if we change to another level:

```{r}
z[1] <- "1"
z
```

--- 

## NAs

* NA stands for not available and represents data that are *missing*. 

* Data analysts have to deal with NAs often.

* In R there is a different kind of NA for each of the basic vector data types.

* There is also the concept of NULL, which represents a zero length list and is often returned by functions or expressions that do not have a specified return value 

--- 

## NAs

* dslabs includes an example dataset with NAs

```{r}
library(dslabs)
na_example[1:20]
```

* The `is.na` function is key for dealing with NAs

```{r}
is.na(na_example[1])
is.na(na_example[17])
is.na(NA)
is.na("NA")
```



--- 

## NAs{.smaller}

* *Caution* logical operators like  and (`&`) and or (`|`) coerce their arguments when needed and possible
* the logical operators evaluate arguments in a "lazy" fashion, and left to right

```{r}
TRUE & NA
TRUE & 0
TRUE | NA
```

--- 

## NaNs and Inf{.smaller}

* A related constant is `NaN` which stands for Not a Number

* `NaN` is a double, coercing it to integer yields an NA

* Inf and -Inf represent values of infinity and minus infinity (RStudio makes using these really annoying)


```{r}
#| code-line-numbers: false

0/0
class(0/0)
sqrt(-1)
log(-1)
1/Inf
Inf-Inf
```

--- 


## Lists

* Data frames are a type of list. 

* Lists permit components of different types and, unlike data frames, different lengths:

```{r}
x <- list(name = "John", id = 112, grades = c(95, 87, 92))
```

* The JSON format is best represented as list in R.

--- 

## Lists

* You can access components in different ways:

```{r}
x$name
x[[1]]
x[["name"]]
```

## Matrices

* Matrices are another widely used data type. 

* They are similar to data frames except all entries need to be of the same type.

* We will learn more about matrices in the High Dimensional Data Analysis part of the class.

---

## Functions {.smaller}

* You can define your own function. The form is like this:

```{r}
#| eval: false
f <- function(x, y, z = 0){
  ### do calculations with x, y, z to compute object
  return(object)
}
```

* the values you pass to the function are called the arguments and they can have default values (e.g above z is 0 unless provided)

* arguments are matched by either name (which takes precedence) or position

* the value returned by a function is either the value specified in a call to `return` or the value of the last statement evaluated
---

## Functions
 
* Here is an example of a function that sums $1,2,\dots,n$

* within the body of a function the arguments are referred to by their symbols and they take the value supplied at the time of invocation

* any symbol found in the body of the function that does not match an argument has to be matched to a value by a process called scoping

```{r}
s <- function(n){
   return(sum(1:n))
}
```

---

## Flow-control and operators

* R has all the standard flow control constructs that most computer langagues do
* if/else; while; repeat; for; break; next
* you can read the manual pages by calling `help` or using `?` (but for the latter you must quote the argument)

```{r}
#| eval: false
 help("for")
 ?"break"
 ?"&"
 ?"/"
```

## Logical Operators

* the short forms `&` and `|` perform element-wise comparisons (vectorized)
* the long forms `&&` and `||` evaluate the first element only, move left to right and return when the result is determined
* errors often occur when a programmer/analyst uses one form, when they want the other (R tries to warn you when it thinks there is a mistake)

## Arithmetic Operators{.smaller}

* these are operators like `^` or `+`
* `?Syntax` will get you the manual page for operator precedence
* when in doubt always use parentheses - it is much clearer to the reader

```{r}

2^1+1
2^(1+1)
TRUE || TRUE && FALSE   # is the same as
TRUE || (TRUE && FALSE) # and different from
(TRUE || TRUE) && FALSE


```

## Functions

* in R functions are *first class* objects - this means they can be passed as arguments, assigned to symbols, stored in other data structures

* in particular they can be passed as arguments to a function and returned as values

* in some languages (e.g. C or Java) functions are not first class objects and they cannot be passed as arguments

* Python uses a fairly similar strategy for functions to the one used in R (as do many other languages)

## Scope{.smaller}

* most of what computer languages do is map symbols (syntax) to values (semantics) and then create an executable program

* when a computer comes upon an expression it parses it and that identifies the symbols that will need to be looked up

```{r parsing}
#| code-line-numbers: false
expr <- parse(text = "a + b * c")
expr[[1]][1]
expr[[1]][2]
expr[[1]][3]
```

* the third part itself is a compound expression that can be decomposed into its parts, `*`, `b` and `c`
* in order to evaluate this the evaluator must find bindings for each of the symbols
* here we expect it wants numbers for `a`, `b` and `c` and functions for `+` and `*`

##Scope

* all computer languages have a set of rules that are used to match these symbols to values
* one commonly used rule is to use lexical scope, but there are lots of different ways this is done

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: false

fun1 = function(x)  x + y
fun1(4)
```

* in `fun1(4)` we probably all agree that the value that should be used for `x` is 4, we probably expect that `+` is a system function

* but what about `y` - where should it's value come from?

## Scope


* Read the function below and try to understand what happens when it is evaluated:

```{r}
f <- function(x){
  cat("y is", y,"\n")
  y <- x
  cat("y is", y,"\n")
  return(y)
}
y <- 2
f(3)

```

---

## Lexical Scope

- lexical scope says that for any function with unbound values you should use the environment at the time the function was created to first look for bindings, in R (and Python) after that you look in the global environment (your workspace) and then in attached packages and system functions.

- why is this useful?

---

## Lexical Scope{.smaller}

* we can write some interesting functions - like a function that evaluates the log likelihood for any given set of data

* here we use `rexp` to generate values from an Exponential distribution

* notice that a function is returned and that R notes that it is a `closure`

```{r mle}
x = rexp(100, rate = 4)

llExp = function(DATA) {
   n = length(DATA)
   sumx = sum(DATA)
   return(function(mu) {n * log(mu) - mu * sumx})
}

myLL = llExp(x)
myLL
```

---

## Lexical Scope Example{.smaller} 


* here we generate different potential values for the rate parameter and plot the log likelihood - the MLE is the maximum of this

```{r}

##possible values for mu
y = seq(3,5,by = 0.1)

plot(y, myLL(y), type="l", xlab="mu", ylab="log likelihood")
```

---


## Name collisions{.smaller}

* what happens when authors of two different packages choose the same name for their functions?

* Look at how this function associated with the symbol `filter` changes in the following code segment:


```{r}
#| eval: false

filter
library(dplyr)
filter
```

* by calling `library(dplyr)` a new package has been put on the search list. You can call `search()` before and after the call to `library`

* one implication of this observation is that users could inadvertently alter computations - and we will want to protect against that

---

## Evaluation and the process used to find bindings{.smaller}

* when evaluating a function R first establishes an *evaluation environment* that contains the formal arguments matched to the supplied arguments

* if the function is in a package, then the Namespace directives of the package are used to augment this evaluation environment

* any symbol not found in that evaluation environment will be searched for in the _Global Environment_ (your workspace).

* And after that the search path (`search()`) in order.

* The evaluator will take the first match it finds and try to use that (sort of - it does know when it is looking for a function)

---

## Namespaces 

* when authoring a package you will want to use Namespaces - the details will not be discussed here

* If a package uses a Namespace then you can explicitly say which `filter` you want using <pkgname>::<functionname>:

```{r}
#| eval: false
stats::filter
dplyr::filter
```

---

## Examples 

* Restart your R Console and study this example:

```{r}
library(dslabs)
exists("murders")
murders <- murders
murders2 <- murders
rm(murders)
exists("murders")
detach("package:dslabs")
exists("murders")
exists("murders2")
```

--- 

## Object Oriented Programming

* R uses object oriented programming (OOP). 

* Base R uses two approaches referred to as S3 and S4, respectively. 

* S3, the original approach, is more common, but has some severe limitations

* The S4 approach is more similar to the conventions used by the Lisp family of languages.

* In S4 there are classes that are used to describe data structures and generic functions, that have methods associated with them

--- 

## Object Oriented Programming

::: {.panel-tabset}

## Time series

```{r}
#| label: co2-plot
#| out-width: "8in"

plot(co2)
```

## Numeric

```{r}
#| label: wrong-co2-plot
#| out-width: "8in"

plot(as.numeric(co2))

```

:::


--- 

## Object Oriented Programming

* Note `co2` is not numeric:

```{r}
class(co2)
```

* The plots are different because `plot` behaves differently with different classes.

--- 

## Object Oriented Programming

* The first `plot` actually calls the function

```{r}
#| eval: false
plot.ts
```

* Notice all the plot functions that start with `plot` by typing `plot.` and then tab.

* The function plot will call different functions depending on the class of the arguments.


## Plots

* Soon we will learn how to use the ggplot2 package to make plots.

* R base does have functions for plotting though

* Some you should know about are:

    - `plot` - mainly for making scatterplots. 
    - `lines` - add lines/curves to an existing plot.
    - `hist` - to make a histogram.
    - `boxplot` - makes boxplots.
    - `image` - uses color to represent entries in a matrix.

---

## Plots

* Although, in general, we recommend using ggplot2, R base plots are often better for quick exploratory plots.

* For example, to make a histogram of values in `x` simply type:

```{r}
#| eval: false
hist(x)
```

* To make a scatter plot of `y` versus `x` and then interpolate we type:

```{r}
#| eval: false
plot(x,y)
lines(x,y)
```

