---
title: "Collaborative analysis of gun legislation time series"
author: "BST 260 students"
---

# Data acquisition

Grab the gun legislation data (originally from [https://www.rand.org/pubs/tools/TLA243-2-v4.html](https://www.rand.org/pubs/tools/TLA243-2-v4.html)).

```{r getd,message=FALSE}
library(readxl)
if (!exists("lawdat")) {
  download.file("https://tinyurl.com/randgunlaw", (tf <- tempfile()))
  lawdat = read_xlsx(tf, 2)
}
```

# Data massaging for ease of use

Rename the variables for convenience.

```{r lk1,message=FALSE}
library(dplyr)
nd = lawdat |> rename( id=`Law ID`, state=State, 
  abbr=`State Postal Abbreviation`, fips=`FIPS Code`, 
  numclass=`Law Class (num)`, lawclass=`Law Class`, lawclasssub=`Law Class Subtype`, 
  handorlong=`Handguns or Long Guns`, effect=`Effect`, 
  changetype=`Type of Change`, edatenote=`Effective Date Note`, 
  effyear=`Effective Date Year`, effmonth=`Effective Date Month`, 
  effday=`Effective Date Day`, citation=`Statutory Citation`, 
  content=`Content`, supyear=`Supersession Date Year`, 
  supmonth=`Supersession Date Month`, supday=`Supersession Date Day`, 
  ageformin=`Age for Minimum Age Laws`, 
  handwaitdays=`Length of Waiting Period (days, handguns)`, 
  addit=`Additional Context and Notes`, 
  discrep=`Discrepancies`)
```

Use lubridate to create `effdate`, a Date-valued column.
```{r dodate,message=FALSE}
library(lubridate)
nd = nd |> mutate(effdate=ymd(paste(effyear, effmonth, effday, sep="-")))
```

# Data visualization

## Static

Visualize the big picture of counts of legislative
events classified as Restrictive or Permissive after 1985.

```{r lkcounts,message=FALSE}
library(ggplot2)
ggplot(nd |> filter(effect=="Restrictive", effdate > ymd("1985-01-01")), 
   aes(x=effdate)) + geom_histogram(binwidth=365) + ylim(0,350) +
   ggtitle("Counts of restrictive legislative changes after 1985")
ggplot(nd |> filter(effect=="Permissive", effdate > ymd("1985-01-01")), 
   aes(x=effdate)) + geom_histogram(binwidth=365) + ylim(0,350) +
   ggtitle("Counts of permissive legislative changes after 1985")
```

## Interactive

Here we use plotly so that we can hover over elements
of the plot to get more information.

```{r doply,message=FALSE}
library(plotly)
g1 = ggplot(nd |> filter(effect=="Permissive", effdate > ymd("1985-01-01")), 
   aes(x=effdate, text=paste(state,lawclass))) + geom_histogram(binwidth=365) +
   ggtitle("Counts of permissive legislative changes after 1985")
ggplotly(g1)
```

# Per-state sections

## Alabama

```{r somecode,echo=FALSE}
# this code written by claude opus 5.5
wrap_hover <- function(x, width = 40, max_lines = Inf, sep = "<br>", escape = TRUE) {
  x <- as.character(x)
  vapply(x, function(s) {
    if (is.na(s) || !nzchar(s)) return(s)
    if (escape) {                       # keep "<", ">", "&" from being read as HTML
      s <- gsub("&", "&amp;", s, fixed = TRUE)
      s <- gsub("<", "&lt;",  s, fixed = TRUE)
      s <- gsub(">", "&gt;",  s, fixed = TRUE)
    }
    # split any word longer than the line width (long IDs, snake_case names, URLs)
    s <- gsub(sprintf("(\\S{%d})(?=\\S)", width - 1), "\\1 ", s, perl = TRUE)
    lines <- strwrap(s, width = width)
    if (length(lines) > max_lines) {
      lines <- lines[seq_len(max_lines)]
      lines[max_lines] <- paste0(lines[max_lines], " \u2026")
    }
    paste(lines, collapse = sep)
  }, character(1), USE.NAMES = FALSE)
}
```

```{r doplyAL}
library(plotly)
g1 = ggplot(nd |> filter(effect=="Permissive", effdate > ymd("1985-01-01"), state=="Alabama"), 
   aes(x=effdate, text=paste(state,lawclass,wrap_hover(content)))) + geom_histogram(binwidth=365) +
   ggtitle("Counts of permissive legislative changes in Alabama after 1985")
ggplotly(g1) |> plotly::style(hoverlabel = list(align = "left"))
```
## Alaska

```{r doplyAK}
library(plotly)
g1 = ggplot(nd |> filter(effect=="Restrictive", effdate > ymd("1985-01-01"), state=="Alaska"), 
   aes(x=effdate, text=paste(state,lawclass,changetype,wrap_hover(content)))) + geom_histogram(binwidth=365) +
   ggtitle("Counts of restrictive legislative changes in Alaska after 1985")
ggplotly(g1) |> plotly::style(hoverlabel = list(align = "left"))
library(plotly)
g1 = ggplot(nd |> filter(effect=="Permissive", effdate > ymd("1985-01-01"), state=="Alaska"), 
   aes(x=effdate, text=paste(state,lawclass,changetype,wrap_hover(content)))) + geom_histogram(binwidth=365) +
   ggtitle("Counts of permissive legislative changes in Alaska after 1985")
ggplotly(g1) |> plotly::style(hoverlabel = list(align = "left"))
```
## Arizona
## Arkansas
## California

```{r doplyCA}
library(plotly)
g1 = ggplot(nd |> filter(effect=="Restrictive", effdate > ymd("1985-01-01"), state=="California"), 
   aes(x=effdate, text=paste(state,lawclass,changetype,wrap_hover(content)))) + geom_histogram(binwidth=365) +
   ggtitle("Counts of restrictive legislative changes in California after 1985")
ggplotly(g1) |> plotly::style(hoverlabel = list(align = "left"))
g1 = ggplot(nd |> filter(effect=="Permissive", effdate > ymd("1985-01-01"), state=="California"), 
   aes(x=effdate, text=paste(state,lawclass,changetype,wrap_hover(content)))) + geom_histogram(binwidth=365) +
   ggtitle("Counts of permissive legislative changes in California after 1985")
ggplotly(g1) |> plotly::style(hoverlabel = list(align = "left"))
```
## Colorado
## Connecticut
## Delaware
## Florida
## Georgia
## Hawaii
## Idaho
## Illinois
## Indiana
## Iowa
## Kansas
## Kentucky
## Louisiana
## Maine
## Maryland
## Massachusetts
## Michigan
## Minnesota
## Mississippi
## Missouri
## Montana
## Nebraska
## Nevada
## New Hampshire
## New Jersey
## New Mexico
## New York
## North Carolina
## North Dakota
## Ohio
## Oklahoma
## Oregon
## Pennsylvania
## Rhode Island
## South Carolina
## South Dakota
## Tennessee
## Texas
## Utah
## Vermont
## Virginia
## Washington
## West Virginia
## Wisconsin
## Wyoming
