---
title: "Association Not Causation"
keywords: "Other topics"
date: "2026-11-09"
format:
  revealjs:
    theme: night
execute:
  echo: true
  fig-align: center
---

```{r setup, include=FALSE}
options(digits = 3)
```

## Association is not causation

- *Association is not causation* is perhaps the most important lesson one can learn in a statistics class.

- *Correlation is not causation* is another way to say this.

- Throughout the statistics part of the book, we have described tools useful for quantifying associations between variables.

- However, we must be careful not to over-interpret these associations.

## Minimal requirements

- **Empirical association** (i.e. Correlation): A relationship must exist between the cause and effect

- **Temporal priority:** The cause must happen before the effect. This is considered the most widely accepted requirement for a causal relationship.

- **There is a dose-response relationship:** increasing the "dose" of the cause increases the effect.

- **Plausibility:** There is a plausible mechanism that explains how the cause leads to the effect. 

- **Exploration and Elimination of alternative causes:** Find confounders and understand their role.


## Association is not causation

- There are many reasons that a variable $X$ can be correlated with a variable $Y$, without having any direct effect on $Y$.

- Today we describe four: spurious correlation, data dredging, reverse causation, and confounders.

- With reverse causation the correlation is real, it is just not causal

## Spurious correlation


```{r divorce-versus-margarine, echo=FALSE, message=FALSE, warning=FALSE, cache=FALSE} 
library(tidyverse) 
library(broom) 
library(dslabs) 
the_title <- paste("Correlation =",  
                round(with(divorce_margarine,  
                           cor(margarine_consumption_per_capita, divorce_rate_maine)),2)) 
divorce_margarine |>  
  ggplot(aes(margarine_consumption_per_capita, divorce_rate_maine)) +  
  geom_point(cex = 3) +  
  geom_smooth(method = "lm") +  
  ggtitle(the_title) + 
  xlab("Margarine Consumption per Capita (lbs)") +  
  ylab("Divorce rate in Maine (per 1000)") 
``` 


## Spurious Associations{.smaller}

- any two things that change in a monotone fashion over time will be correlated
    - this can also be caused, as we will see by confounding (where time is often an unadjusted confounder)

- if you look at enough tests, one will be positive (data dredging) 
    - you should always ask: "How many tests did you do?"
    - use of the False Discovery Rate (FDR) and other tools reduces this issue

- outlying observations can cause high correlation 
    - outliers are often spurious but not always


## Spurious correlation


```{r} 
library(tidyverse) 
N <- 25 
g <- 1000000 
sim_data <- tibble(group = rep(1:g, each = N),  
                   x = rnorm(N*g),  
                   y = rnorm(N*g)) 
``` 


## Spurious correlation

```{r} 
res <- sim_data |>  
  group_by(group) |>  
  summarize(r = cor(x, y)) |>  
  arrange(desc(r)) 
res 
``` 

- We see a maximum correlation of `r max(res$r)`.

## Spurious correlation

```{r dredging, eval=FALSE} 
sim_data |> filter(group == res$group[which.max(res$r)]) |> 
  ggplot(aes(x, y)) + 
  geom_point() +  
  geom_smooth(method = "lm") 
``` 


## Spurious correlation

```{r dredging-run, echo=FALSE} 
sim_data |> filter(group == res$group[which.max(res$r)]) |> 
  ggplot(aes(x, y)) + 
  geom_point() +  
  geom_smooth(method = "lm") 
``` 


## Spurious correlation

- Sample correlation is a random variable:


```{r null-corr-hist-run, echo=FALSE} 
res |> ggplot(aes(x=r)) + geom_histogram(binwidth = 0.1, color = "black") 
``` 


## Spurious correlation

- It's simply a mathematical fact that if we observe `r cat(prettyNum(g, big.mark=",",scientific=FALSE))` random correlations that are expected to be 0, but have a standard error of `r sd(res$r)`, then the largest one will be close to 1.

## Spurious correlation


```{r, message=FALSE, warning=FALSE} 
library(broom) 
sim_data |>  
  filter(group == res$group[which.max(res$r)]) |> 
  reframe(tidy(lm(y ~ x))) |>  
  filter(term == "x") 
``` 



## Spurious correlation

- This particular form of data dredging is referred to as *p-hacking*.

- P-hacking is a topic of much discussion because it poses a problem in scientific publications.

- Since publishers tend to reward statistically significant results over negative results, there is an incentive to report significant results.

## Spurious correlation

- In epidemiology and the social sciences, for example, researchers may look for associations between an adverse outcome and several exposures, and report only the one exposure that resulted in a small p-value.

- Similarly with high throughput biology it is routine to test 100's to 1000's of genes, genetic variants, epigenetic sites

- unless care is taken to perform some form of p-value adjustment these results are unlikely to replicate


## Spurious correlation

- Furthermore, they might try fitting several different models to account for confounding and choose the one that yields the smallest p-value.

## Spurious correlation

- In experimental disciplines, an experiment might be repeated more than once, yet only the results of the one experiment with a small p-value reported.

## Spurious correlation

- This does not necessarily happen due to unethical behavior, but rather as a result of statistical ignorance or wishful thinking.

- In advanced statistics courses, you can learn methods to adjust for these multiple comparisons.



## Outliers


```{r} 
set.seed(1985) 
x <- rnorm(100,100,1) 
y <- rnorm(100,84,1) 
x[-23] <- scale(x[-23]) 
y[-23] <- scale(y[-23]) 
``` 

## Outliers

```{r outlier-run, echo=FALSE} 
qplot(x, y) 
``` 


## Outliers


```{r} 
cor(x,y) 
``` 

This high correlation is driven by the one outlier.

## Outliers

- If we remove this outlier, the correlation is greatly reduced to almost 0:

```{r} 
cor(x[-23], y[-23]) 
``` 


## Reversing cause and effect

- Another way association is confused with causation is when the cause and effect are reversed.

- An example of this is claiming that tutoring makes students perform worse because they test lower than peers that are not tutored.

- In this case, the tutoring is not causing the low test scores, but the other way around.

## Reversing cause and effect

Quote from [NY Times](https://opinionator.blogs.nytimes.com/2014/04/12/parental-involvement-is-overrated):


>> When we examined whether regular help with homework had a positive impact on children's academic performance, we were quite startled by what we found. Regardless of a family's social class, racial or ethnic background, or a child's grade level, consistent homework help almost never improved test scores or grades... 

- A more likely possibility is that the children needing regular parental help, receive this help because they don't perform well in school.

## Reversing cause and effect


- If we fit the model:

$$
X_i = \beta_0 + \beta_1 y_i + \varepsilon_i, i=1, \dots, N
$$ 

- where $X_i$ is the father height and $y_i$ is the son height, we do get a statistically significant result.


```{r, echo=FALSE, cache=FALSE} 
library(HistData) 
set.seed(1983) 
galton_heights <- GaltonFamilies |> 
  filter(gender == "male") |> 
  group_by(family) |> 
  sample_n(1) |> 
  ungroup() |> 
  select(father, childHeight) |> 
  rename(son = childHeight) 
``` 

```{r} 
galton_heights |> reframe(tidy(lm(father ~ son))) 
``` 


## Reversing cause and effect

- The model fits the data very well.

- The model is technically correct.

- The estimates and p-values were obtained correctly as well.

- But we cannot interpret this as saying that the heights of the son's has any causal effect on the height of the fathers


## Confounders

- Confounders are a common reason for associations being misinterpreted.

- When studying the relationship between and exposure and an outcome a confounder is third variable that:

     - Is related to the exposure 

     - Affects the outcome, and is not caused by the exposure

- hence the confounder can affect the perceived relationship between the exposure and the outcome.


## Confounders

- Earlier, when studying baseball data, we saw how Home Runs were a confounder that resulted in a higher correlation than expected when studying the relationship between Bases on Balls and Runs.

- In some cases, we can use linear models to account for confounders.

- However, this is not always the case.

## Confounders

- Incorrect interpretation due to confounders is ubiquitous in the lay press and they are often hard to detect.

- Here, we present a widely used example related to college admissions.

## UC Berkeley admissions

```{r} 
two_by_two <- admissions |> group_by(gender) |>  
  summarize(total_admitted = round(sum(admitted / 100 * applicants)),  
            not_admitted = sum(applicants) - sum(total_admitted)) 
two_by_two |> 
  mutate(percent = total_admitted/(total_admitted + not_admitted)*100)
```

```{r}
two_by_two <- select(two_by_two, -gender)
chisq.test(two_by_two)$p.value 
``` 

## UC Berkeley admissions

- Closer inspection shows a paradoxical result:

```{r} 
admissions |> select(major, gender, admitted) |> 
  pivot_wider(names_from = "gender", values_from = "admitted") |> 
  mutate(women_minus_men = women - men) 
``` 


## UC Berkeley admissions

- What's going on?

- This actually can happen if an uncounted confounder is driving most of the variability.

- This particular version is called *Simpson's Paradox*

## UC Berkeley admissions

```{r uc-berkeley-majors-run, echo=FALSE} 
admissions |>  
  group_by(major) |>  
  summarize(major_selectivity = sum(admitted * applicants)/sum(applicants), 
            percent_women_applicants = sum(applicants * (gender=="women")) / 
                                             sum(applicants) * 100) |> 
  ggplot(aes(major_selectivity, percent_women_applicants, label = major)) + 
  geom_text() 
``` 



## Confounding explained 

```{r confounding, echo=FALSE} 
admissions |> 
  mutate(yes = round(admitted/100*applicants), no = applicants - yes) |> 
  select(-applicants, -admitted) |> 
  gather(admission, number_of_students, -c("major", "gender")) |> 
  ggplot(aes(gender, number_of_students, fill = admission)) + 
  geom_bar(stat = "identity", position = "stack") + 
  facet_wrap(. ~ major) 
``` 



## Average after stratifying

```{r admission-by-major-run, echo=FALSE} 
admissions |>  
  ggplot(aes(major, admitted, col = gender, size = applicants)) + 
  geom_point() 
``` 

## Average after stratifying


- If we average the difference by major, we find that the percent is actually 3.5% higher for women.

```{r} 
admissions |>  group_by(gender) |> 
  summarize(average = mean(admitted)) 
``` 



## Simpson's paradox

- The case we have just covered is an example of Simpson's paradox.

- It is called a paradox because we see the sign of the correlation flip when comparing the entire population to specific strata.



## Simpson's paradox


- Simulated $X$, $Y$, and $Z$:

```{r simpsons-paradox, echo=FALSE} 
N <- 100 
Sigma <- matrix(c(1,0.75,0.75, 1), 2, 2)*1.5 
means <- list(c(x = 11, y = 3),  
              c(x = 9, y = 5),  
              c(x = 7, y = 7),  
              c(x = 5, y = 9),  
              c(x = 3, y = 11)) 
dat <- lapply(means, function(mu){ 
  res <- MASS::mvrnorm(N, mu, Sigma) 
  colnames(res) <- c("x", "y") 
  res 
}) 
dat <- do.call(rbind, dat) |>  
  as_tibble() |> 
  mutate(z = as.character(rep(seq_along(means), each = N))) 
dat |> ggplot(aes(x, y)) + geom_point(alpha = 0.5) + 
  ggtitle(paste("Correlation = ", round(cor(dat$x, dat$y), 2))) 
``` 


## Simpson's paradox

- You can see that $X$ and $Y$ are negatively correlated.

- However, once we stratify by $Z$ (shown in different colors below), another pattern emerges.


## Simpson's paradox

```{r}
#| echo: false
dat |> ggplot(aes(x, y)) + geom_point(alpha = 0.5) + 
  ggtitle(paste("Correlation = ", round(cor(dat$x, dat$y), 2))) 

```

## Simpson's paradox

```{r simpsons-paradox-explained, echo=FALSE} 
means <- do.call(rbind, means) |>  
  as_tibble() |> 
  mutate(z = as.character(seq_along(means))) 
corrs <- dat |> group_by(z) |> summarize(cor = cor(x, y)) |> pull(cor) 
dat |> ggplot(aes(x, y, color = z)) +  
  geom_point(show.legend = FALSE, alpha = 0.5) + 
  ggtitle(paste("Correlations =",  paste(signif(corrs,2), collapse=" "))) + 
  annotate("text", x = means$x, y = means$y, label = paste("z =", means$z), cex = 5)   
``` 


## Simpson's paradox

- It is really $Z$ that is negatively correlated with $X$.

- If we stratify by $Z$, the $X$ and $Y$ are actually positively correlated, as seen in the plot above.

## Collider Bias

- when studying medical records an interesting phenomenon called collider bias can occur

- Example from Wikipedia: https://en.wikipedia.org/wiki/Berkson%27s_paradox

An example presented by mathematician Jordan Ellenberg is that of a dating pool, measured on axes of niceness and handsomeness. A person might conclude from their own dating experience that "the handsome ones tend not to be nice, and the nice ones tend not to be handsome".

## Collider Bias Example

- Alice decides to only date men that that are above some line of handsomeness and niceness

- But as a result the really handsome ones do not have to be that nice and the really nice ones do not need to be that handsome

- so now Alice notices a correlation in her dating pool even when there is no such correlation in the population

## Medically relevant examples

- studies of hospital records can come to the wrong conclusions

- being in hospital requires that the person have something serious and medically adverse

- unless care is taken spurious negative associations can be found

- when looking at everyone in the hospital for appendicitis the researchers find that few if any of them have kidney disease

## Medical Example

- and when comparing their appendicitis patients to controls, taken from patients in the hospital they find this association to be significant

- it would be wrong to believe that this suggests that kidney disease is *protective* against appendicitis

- they have to have something wrong with them, and since both kidney disease and appendicitis are rare it is very unlikely that many patients will have both

