---
title: Problem set 4
date: 2026-10-09
format: html
---

# Doing problem set 4

Retrieve this quarto document and place a copy in your bst260 repository.

You can find a copy at [this URL](https://tinyurl.com/260pset04).

Call it pset04.qmd.

You will be editing the document directly to solve exercise 6.  

Place your
name as the author of the solutions in the yaml header above.  

When
you have finished editing this document with your solutions
place it and your compiled html in the solutions folder of your bst260
repository.

# Descriptive statistics for the NHANES dataset

The NHANES package on CRAN includes a dataset called "NHANES"
that addresses issues of sampling to permit direct statistical analyses.

Example analyses of the weighted survey data assembled by
the National Center for Health Statistics are available.  An
example is [here](https://wwwn.cdc.gov/nchs/data/Tutorials/Code/DB303_R.r);
also of interest is the nhanesA package, which leverages
a containerized relational database.  A publication is [available](https://pmc.ncbi.nlm.nih.gov/articles/PMC11020206/).

Here we use the 'resampled' version for simplicity.

## Software setup

Let's get the necessary software.  Use `install.packages` for
any packages you are missing.

```{r lknh, message=FALSE}
library(NHANES)
data(NHANES)
library(dplyr)
library(table1)
library(ggplot2)
nh11 = NHANES |> filter(SurveyYr == "2011_12")
```

## "Table 1"

The `table1` package/function provides a powerful
approach to producing summary statistics. We will focus
on the measured quantity "diastolic blood pressure" as
obtained by averaging several repeat measures for each
participant visit.

```{r femd, message=FALSE}
#| tbl-cap: "Basic DBP statistics for females"
#| results: asis
table1(~BPDiaAve|Race3, data=nh11 |> filter(Sex=="female"))
```

```{r mald, message=FALSE}
#| tbl-cap: "Basic DBP statistics for males"
#| results: asis
table1(~BPDiaAve|Race3, data=nh11 |> filter(Sex=="male"))
```

### Exercise 1.  

Write and run the _dplyr_ code to reproduce the means,
standard deviations and number of missing values for BPDiaAve in each table.
The result should have column names "Sex", "Race3", "mean", "sd", "nmiss".
Records for females should come first and rows should be sorted by race
within gender.


### Exercise 2.

Something very implausible is present in the data.  Describe it and reproduce
the table1 results and dplyr summaries after "fixing" it.

### Exercise 3: Compact estimation procedures for averages

R's built-in modeling functions can save a lot of
filtering and summarization activity.  To estimate
average diastolic blood pressures for combinations of race and gender, we can use

```{r dolm1}
lm1 = lm(BPDiaAve~Race3*Sex-1, data=nh11, na.action=na.omit)
lm1$coef
```
The syntax `y ~ a * b - 1` for numerical `y` and factor variables `a` and `b`
is described as fitting "first-order interactions".
Subgroups of individuals are formed for
every combination of factor levels, and the mean value of `y`
is computed for each subgroup.  The `-1` suppresses an
intercept term.

R uses a peculiar naming convention for the estimated parameters.
`Race3Asian` is used to show that the factor level for Race3 is Asian.
`Sexmale` has a coefficient.  This indicates to us that the 'reference'
sex is female.  Finally colon is used for combinations of
factor values.  We can recover the mean value for
Black males by adding the coefficients for Race3Black,
Sexmale, and Race3Black:Sexmale.

Create the compact estimates using the version of the data
"repaired" in Exercise 2 and verify that the estimates
from lm recover the dplyr summaries.

### Exercise 4: Visualizing DBP distributions by age and race

Here is classic R list manipulation and plotting code:
```{r lkmmm}
par(mfrow=c(3,2),las=2,mar=c(5,4,2,2))
ss = split(nh11, nh11$Race3)
tt = lapply(names(ss), function(x) {d = ss[[x]]; 
   boxplot(split(d$BPDiaAve, d$AgeDecade), main=x, ylim=c(0,120))})
```

Write and run the ggplot2 code to produce a similar display.

### Exercise 5: Do ethnic groups have comparable age distributions?


Use the compact estimation procedure with Age as the "y"
variable.  What do you see as the most prominent indications
of heterogeneity in age distributions across the different
group combinations?

Describe how to form observed and expected counts in the
cells of the table
```{r lktab}
with(nh11, table(AgeDecade, Race3))
```

### Exercise 6: Cards to Dice

Modify all the quarto and R following to replace
draws from a deck of shuffled
cards with tosses of a single six-faced die.

As an example, you might start with code like
```
die = c(1,2,3,4,5,6)
roll = function(die) sample(die,size=1)
set.seed(1234)
table(replicate(100, roll(die)))
```
to model 100 tosses of a fair die.

## Modeling cards and shuffled decks

### Decks as vectors

We followed online code to produce
```{r do1}
suits <- c("Spades", "Hearts", "Diamonds", "Clubs")
ranks <- c("Ace", 2:10, "Jack", "Queen", "King")
# Combine ranks and suits to make 52 unique cards
deck <- paste(rep(ranks, 4), rep(suits, each = 13))
```
A deck is a vector with 52 elements.
This vector "models" a deck of playing cards.

## Draws as samples

With R's `sample` function, we can model shuffling.
We use `set.seed` to obtain a reproducible random
number generation state.

```{r lkshuf}
top_card = function(x) { 
   stopifnot(length(x)==52, is.atomic(x))
   sample(x, size=1)
}
set.seed(1234)
drs = replicate(2000, top_card(deck))
head(table(drs))
mean(table(drs))
2000/52
```

## A model for fairness

### Intuitions

For a fair deck with thorough shuffling and replacement, we 
have an intuition that each
card type has the same chance of being drawn as the top card.

In 2000 draws, the _expected_ number of appearances of "2 Hearts"
is 2000/52, and that is the same for all card types.

### Formalism

Formally, we can abstract from the actual names of the cards and
represent the shuffling/drawing procedure in this way:

Let $i = 1, \ldots, 52$ denote card types

Let $j = 1, \ldots, 2000$ denote shuffle/draw/replace events

Define

$$
C_{ij} =
\begin{cases}
  1 & \text{if draw } j \text{ yields card type } i, \\
  0 & \text{otherwise.}
\end{cases}
$$

Now define $X_i = \sum_j C_{ij}$.  Our statement above about the
intuitive expectation on the number of appearances of any given
card type will be written
$$
EX_i = 2000/52 \text{ for all } i = 1,\ldots, 52.
$$

This is equivalent to the following formulation.  Let $p_i, i=1, \ldots, 52$
denote the _probability_ (not yet defined in any formal way) that our shuffle/draw procedure
yields card of type $i$.  Then $EX_i = 2000 p_i$ for all $i$.  For the
general case of $n$ draws, $EX_i = n p_i$.

## Measuring departure from fairness

A well-established metric for assessing departure from fairness is
called "chi-squared".  We will use the symbol $D$ to represent this
measure.  Large values of $D$ can be indicative of departure of the
observed frequencies from the expected values.

$$
D = \sum_{i=1}^{52} \frac{(X_i - EX_i)^2}{EX_i}
$$

We'll implement $D$ as `mystat` and show that it is equivalent to
a quantity computed in R with `chisq.test`.

```{r OmE}
mystat = function(dat, expected) sum((table(dat)-expected)^2/expected)
ExpectedCount = 2000/52
sum((table(drs)-ExpectedCount)^2/ExpectedCount)
chisq.test(table(drs))
mystat(drs, ExpectedCount)
```

Note that `chisq.test` reports the same statistic as `mystat`, but
includes values for `df` (degrees of freedom) and `p-value`.  "Degrees of freedom"
refers here to the fact that we have a fixed number of count categories (52 card types),
and if 51 are known, the value for the 52nd can be deduced -- it is "unfree".

The p-value is a quantity that, when small, can be indicative of
a departure from the expected outcome of an experiment.  We will
return to that concept at a later date.

## Replicating our test of fairness many times

Any given collection of shuffle/draw events will yield
a value of $D$.  Different series will yield different
values of $D$, but when the deck is fair, the variation
in $D$ is prescribed by the mathematics of probability.

We will make a histogram of 100 replicates of 2000 draws
to exhibit variation in $D$.

```{r domore}
set.seed(1234)
predist = lapply(1:100, function(x) replicate(2000, top_card(deck)))
dist = sapply(predist, function(x)mystat(x, expected=2000/52))
hist(dist, prob=TRUE)
```

With 1000 replicates, we get a more pleasing display.

```{r domore2, cache=TRUE}
set.seed(1234)
predist = lapply(1:1000, function(x) replicate(2000, top_card(deck)))
dist = sapply(predist, function(x)mystat(x, expected=2000/52))
hist(dist, probability=TRUE)
```

The form of this histogram (drawn in R with the `probability` argument set to TRUE)
can be approximated by the density of the chi-squared distribution
with 51 degrees of freedom.

```{r lkdens}
hist(dist, probability=TRUE)
lines(1:100, dchisq(1:100, 51))
```

The "51" fixes the degrees of freedom of the model for which
the density will be drawn.

## Interpret

What's the difference between these two runs?

```{r lkbad}
set.seed(1234)
baddeck = deck
baddeck[1] = baddeck[52]
newdrs = replicate(2000, top_card(baddeck))
mystat(newdrs, ExpectedCount)
```

```{r lkbad2}
set.seed(1234)
baddeck[2] = baddeck[51]
newdrs = replicate(2000, top_card(baddeck))
mystat(newdrs, ExpectedCount)
```

